Follow-up Samples

### 2017-02-09 submission/manuscript_outline_20161024.Rmd
discussion improvements pt1

***Bacterial Community and FIT***
We first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma or carcinoma.  What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. We also found that there were larger changes fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The broad shift in bacterial community structure before and after surgery was visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma, or carcinoma for any metric [Table S1].  We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lession, adenoma only, or carcinoma only [Figure S1].

***Previously Associated Cancer Bacteria***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a difference in initial and follow up samples for lesion and carcinoma for relative abundance of *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`). In contrast, there was no differences relative abundance in any of these OTUs for those with adenoma [Figure 2]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of these OTUs [Figure 2].

***Full and Reduced Model***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma.  In order to investigate this we created two different models: one to classify lesion versus normal and one to classify pre versus post samples based on the bacterial community and FIT.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while the initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  By identifying the most important variables for each respective model and then reducing them to only these factors we were able to increase the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).        
The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].  The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].  Interestingly, the test set AUC range for the initial sample model performed much better then the training set AUCs.  Both the full and reduced initial sample models overlapped with each other [Figure 3b] but there was a marked decrease in the ROC curve for the final model used.

@@

***Treatment and Time Differences***
After observing these changes in positive probability We wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the before sample model was similar there was a significant difference in decreased positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).  Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.  Within our study there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).

@@

## Discussion
There was no difference in alpha diversity metrics between the initial and follow up samples [Table S1].  Yet based on thetayc there was a significant difference between initial and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.  The change in FIT was also different between initial and follow up for those with adenoma or carcinoma [Figure 1b].  There was also a significant difference in the overall bacterial community structure for those in the carcinoma group [Figure 1d] before and after surgery but not for the adenoma group [Figure 1c].  Investigation of the OTU relative abundance before and after surgery found no single OTU to be significantly different [Figure S1].

>>>>

***Bacterial Community and FIT***
We first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma or carcinoma.  What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The broad shift in bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma, or carcinoma for any metric tested [Table S1].  We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lesion, adenoma, or carcinoma [Figure S1].

***Previously Associated Cancer Bacteria***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a difference in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`). In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of these respective OTUs [Figure 2].

***Full and Reduced Model***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma.  In order to investigate this we created two different models: one to classify lesion versus normal and one to classify pre versus post samples based on the bacterial community and FIT.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while the initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  By identifying the most important variables for each respective model and then reducing them to only these factors we were able to increase the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).        
The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap with each other [Figure 3a].  The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].  Interestingly, the test set AUC range for the initial sample model performed much better then the training set AUCs.  Both the full and reduced initial sample models overlapped with each other [Figure 3b] but there was a marked decrease in the ROC curve for the final before sample model used.

@@

***Treatment and Time Differences***
After observing these changes in positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the before sample model was similar there was a significant difference in decreased positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).  Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.  Within our study there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).

@@

## Discussion
From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There was much larger difference between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for indivdiuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, oberseved OTUs, or evenness regardless of whether the indidivual had an adenoma or carcinoma [Table S1].  There was alos no differences in relative abundnace of any specific OTU for lesion, adenoma, or carcinoma [Figure s1].
